---
title: "The Wisdom and Madness of Crowds"
subtitle: "RIKEN Wako Science Camp 2026 — Day 3"
author: "Computational Group Dynamics (COGNAC) Collaboration Unit, RIKEN CBS"
format:
  revealjs:
    theme: [default]
    slide-number: true
    embed-resources: true
    transition: slide
execute:
  echo: false
  warning: false
  message: false
---

## Yesterday {.center}

You built an agent that **learns alone**.

Today: agents that **watch each other**.

Spoiler: societies appear — with all their glory and their bugs.

## Why copy others?

New town, hungry. One ramen shop has a queue, one is empty.

::: {.fragment}
Other people's choices carry **information**.

Copying is often *smart* — humans are champions of
**social learning**（社会学習）. It gave us culture,
language, science... and TikTok.
:::

## Two knobs of a social learner

$$p(\text{choice}) = (1-\sigma)\, p_{\text{own experience}} + \sigma\, p_{\text{copy others}}$$

- $\sigma$ (sigma): **how much** to rely on others (0 → 1)
- $\theta$ (theta): **how strongly** to follow majorities
  - $\theta = 1$: copy in proportion
  - $\theta > 1$: **conformity** — exaggerate the majority

::: {.fragment}
Different values = different *social strategies*.
Which strategy is the smartest?
:::

## 1906: the ox and the crowd

At a fair, ~800 people guessed the weight of an ox.

Nearly everyone was wrong...

::: {.fragment}
...but the **average of all guesses** was off by **less than 1%**.

**Wisdom of crowds**: independent errors point in different
directions and *cancel out*.
:::

## Copying makes a group smarter...

```{r}
#| fig-height: 4.5
source("R/bandit_functions.R")
library(ggplot2)
set.seed(7)
mix <- do.call(rbind, lapply(c(0, 0.6), function(s) {
  perf <- average_performance(n_reps = 25, n_agents = 10, n_trials = 100,
                              sigma = s, theta = 2)
  transform(perf, society = ifelse(s == 0, "no copying (sigma=0)", "copying (sigma=0.6)"))
}))
ggplot(mix, aes(trial, correct, color = society)) +
  geom_line(linewidth = 1.2) + ylim(0, 1) +
  labs(y = "fraction choosing better option") +
  theme_minimal(base_size = 18) + theme(legend.position = "bottom")
```

Good discoveries spread; individual mistakes get filtered out.

## ...until the world changes

```{r}
#| fig-height: 4.5
set.seed(11)
sweep <- do.call(rbind, lapply(c(0, 0.9), function(s) {
  perf <- average_performance(n_reps = 25, n_agents = 10, n_trials = 150,
                              sigma = s, theta = 2, change_point = 76)
  transform(perf, society = ifelse(s == 0, "individual learners", "heavy copiers (sigma=0.9)"))
}))
ggplot(sweep, aes(trial, correct, color = society)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 75.5, linetype = "dashed") +
  ylim(0, 1) +
  labs(y = "fraction choosing (currently) best option",
       caption = "dashed line: the best option secretly swaps") +
  theme_minimal(base_size = 18) + theme(legend.position = "bottom")
```

## Herding（群衆行動）

After the change, heavy copiers keep choosing the **old** best option.

Everyone copies everyone; the majority feeds on itself.

::: {.incremental}
- fashion bubbles 👟
- stock market crashes 📉
- misinformation cascades 📱
:::

::: {.fragment}
**Same strategy. Different environment. Opposite result.**
:::

## This is real science, done here

Real people played this game in groups, online
(Toyokawa, Whalen & Laland 2019, *Nature Human Behaviour*):

- moderate copying → collective intelligence ✅
- strong conformity + changing world → herding ❌

::: {.fragment}
Today, **you** run this research program —
with your own questions.
:::

## Your mission (until tomorrow)

1. **Play the game yourselves** — `R/play_bandit_game.R`
   (solo, then with social hints). You are the data! 📊
2. **Pick a question** from `project/mini_project_menu.md`
   (or invent your own)
3. **Simulate, plot, interpret**
4. Tonight: build slides. Tomorrow: **present like scientists**

## A good 10-minute talk

1. The question — *why is it interesting?* (2 min)
2. The model — *the rules of your toy world* (2 min)
3. What you did (1 min)
4. Results — *1–2 clear figures* (3 min)
5. What it means + what you'd do next (2 min)

::: {.fragment}
Template ready for you: `slides/final_presentation_template.qmd`

**Go! 🚀**
:::
