---
title: "Social Learning, Crowds & Your Project"
subtitle: "RIKEN Wako Science Camp 2026 — Day 3"
author: "Computational Group Dynamics (COGNAC) Collaboration Unit, RIKEN CBS"
format:
  revealjs:
    theme: [default]
    slide-number: true
    embed-resources: true
    transition: slide
execute:
  echo: false
  warning: false
  message: false
---

## Today {.center}

**Linked hands-on notebooks（演習用ノートブック）:**<br>
`tutorials/04_social_learning_notebook.Rmd`<br>
`tutorials/05_collective_behaviour_notebook.Rmd`

You built an agent that **learns alone**.

Today: agents that **watch each other**.

Societies appear. They can be wise, or they can make mistakes.

## Why copy others?

New town, hungry. One ramen shop has a queue, one is empty.

::: {.fragment}
Other people's choices carry **information**.

Copying is often *smart* — humans are champions of
**social learning**（社会学習）. It gave us culture,
language, science... and TikTok.
:::

## Two knobs of a social learner {.smaller}

$$
p(\text{choice})
= (1-\sigma)\,p_{\text{individual}}
+ \sigma\,p_{\text{social}}
$$

- $p_{\text{individual}}$: your own experience
- $p_{\text{social}}$: other people's choices
- $\sigma$: **how much** to rely on others (0 → 1)
$$
p_{\text{social}} \propto \text{count}^{\theta}
$$

- $\theta$: **how strongly** to follow majorities
  - $\theta = 1$: copy in proportion
  - $\theta > 1$: **conformity**（同調）— exaggerate the majority


::: {.fragment}
Different values = different *social strategies*.
Which strategy is the smartest?
:::

## Hands-on 1: Tutorial 4 {.center}

Open:

`tutorials/04_social_learning_notebook.Rmd`

Build a society of learners. Change $\sigma$ (sigma) and $\theta$ (theta).

**Predict first. Then run.**

<!-- Resume this deck after Tutorial 4. -->

## From learners to crowds {.center}

You now have agents that can copy.

What happens to the **whole group**?

## 1906: the ox and the crowd

At a fair, about 800 people guessed the weight of an ox.

Nearly everyone was wrong...

::: {.fragment}
...but the **average of all guesses** was off by **less than 1%**.

**Wisdom of crowds**: independent errors point in different
directions and *cancel out*.
:::

## Copying makes a group smarter...

```{r}
#| fig-height: 4.5
source("R/bandit_functions.R")
library(ggplot2)
set.seed(7)
mix <- do.call(rbind, lapply(c(0, 0.6), function(s) {
  perf <- average_performance(n_reps = 25, n_agents = 10, n_trials = 100,
                              reward_probs = c(0.3, 0.5),
                              alpha = 0.3, beta = 5,
                              sigma = s, theta = 2)
  transform(perf, society = ifelse(s == 0, "no copying (sigma=0)", "copying (sigma=0.6)"))
}))
ggplot(mix, aes(trial, correct, color = society)) +
  geom_line(linewidth = 1.2) + ylim(0, 1) +
  labs(y = "fraction choosing better option") +
  theme_minimal(base_size = 18) + theme(legend.position = "bottom")
```

Good discoveries spread; individual mistakes get filtered out.

## What if the world changes? {.center}

The best option secretly swaps halfway through.

Which learners will adapt faster?

Write down your prediction before you see the result.

## Hands-on 2: Tutorial 5 {.center}

Open:

`tutorials/05_collective_behaviour_notebook.Rmd`

Find collective wisdom. Then change the world and look for herding.

**Predict first. Then run.**

<!-- Resume this deck after Tutorial 5. -->

## ...until the world changes

```{r}
#| fig-height: 4.5
set.seed(11)
sweep <- do.call(rbind, lapply(c(0, 0.7), function(s) {
  perf <- average_performance(n_reps = 25, n_agents = 10, n_trials = 150,
                              reward_probs = c(0.5, 0.7),
                              alpha = 0.7, beta = 5,
                              sigma = s, theta = 4, change_point = 76)
  transform(perf, society = ifelse(s == 0, "individual learners", "strong social learners (sigma=0.7)"))
}))
ggplot(sweep, aes(trial, correct, color = society)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 75.5, linetype = "dashed") +
  ylim(0, 1) +
  labs(y = "fraction choosing (currently) best option",
       caption = "dashed line: the best option secretly swaps") +
  theme_minimal(base_size = 18) + theme(legend.position = "bottom")
```

## Herding（群れ行動・付和雷同）

After the change, heavy copiers keep choosing the **old** best option.

Everyone copies everyone; the majority keeps making itself bigger.

::: {.incremental}
- fast-changing fashion trends 👟
- stock market crashes 📉
- misinformation cascades 📱
:::

::: {.fragment}
**Same strategy. Different environment. Opposite result.**
:::

## This is real science, done here

Real people played this game in groups, online
(Toyokawa, Whalen & Laland 2019, *Nature Human Behaviour*):

- moderate copying → collective intelligence ✅
- strong conformity + changing world → herding ❌

::: {.fragment}
Today, **you** run this research program —
with your own questions.
:::

## Your group project

1. **Play the game yourselves** — `R/play_bandit_game.R`
   (solo, then with social hints). You are the data! 📊
2. **Pick a question** from `project/mini_project_menu.md`
   (or invent your own)
3. **Simulate, plot, interpret**
4. **Export your figures before 15:45**
5. **Start your slides before leaving the lab**

## A good 10-minute talk

1. The question — *why is it interesting?* (2 min)
2. The model — *the rules of your toy world* (2 min)
3. Your prediction — *what did you expect?* (1 min)
4. What you did (1 min)
5. Results — *1–2 clear figures* (2 min)
6. What it means + what you'd do next (2 min)

## Time to get started!
Template ready for you: `slides/final_presentation_template.qmd`

Project menu: `project/mini_project_menu.md`

**Choose your question. Predict. Then start! 🚀**
