---
title: 'Tutorial 3: An Agent That Learns (Interactive Notebook)'
subtitle: Slot machines, trial and error, and the explore–exploit dilemma
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    theme: cosmo
editor_options:
  chunk_output_type: inline
---

::: {.callout-note}
## How to use this notebook

Open `ScienceHighSchool.Rproj` first. Run each code block with the green **Run** button, or place your cursor in the code and press **Cmd/Ctrl + Enter**.

Change one small thing at a time. Then run the block again and look at the result.
:::

```{r setup, include=FALSE}
project_root <- normalizePath("..")
if (!file.exists(file.path(project_root, "ScienceHighSchool.Rproj"))) {
  project_root <- getwd()
}
if (!file.exists(file.path(project_root, "ScienceHighSchool.Rproj"))) {
  stop("Please open ScienceHighSchool.Rproj before running this notebook.")
}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = project_root)
```


::: {.callout-note}
## Word list（単語リスト）

| English | 日本語 | Meaning |
|---|---|---|
| agent | エージェント | a (simulated) individual that makes choices |
| reward | 報酬 | the points/food/money an agent receives |
| value estimate | 価値の推定 | the agent's guess about how good an option is |
| learning rate | 学習率 | how quickly the agent updates its guess |
| explore / exploit | 探索・活用 | try something new / stick with what works |
:::

## The slot machine problem

Imagine two slot machines（スロットマシン）:

- Machine A pays a coin **30%** of the time
- Machine B pays a coin **70%** of the time

You do **not** know these numbers. You have 100 turns.
How do you find the good machine — and how do you get rich?

This is called the **multi-armed bandit problem**, and it is one of the
most famous problems in psychology, neuroscience, and AI. New restaurant
or favourite restaurant? New song or favourite playlist? Same problem:
**explore** or **exploit**?

Today we build an **agent**: a small simulated creature that faces this
problem and *learns from experience*.

## Idea 1: keep a value estimate for each option

Our agent keeps a number $Q$ for each machine — its current guess of
"how good is this machine?" Both start at 0 (no idea).

After every play, the agent updates its guess:

$$
Q_{\text{new}} = Q_{\text{old}} + \alpha \times (\text{reward} - Q_{\text{old}})
$$

The part $(\text{reward} - Q_{\text{old}})$ is the **prediction error**:
*"how different was reality from what I expected?"*

- Got a coin ($1$) when you expected little ($Q=0.2$)? Positive surprise → $Q$ goes **up**.
- Got nothing ($0$) when you expected a lot ($Q=0.8$)? Negative surprise → $Q$ goes **down**.

The **learning rate** $\alpha$ (alpha, between 0 and 1) controls how big
each update is: how much the newest experience counts.

```{r}
alpha <- 0.3
Q <- 0                      # current guess for one machine
reward <- 1                 # we just won a coin!
Q <- Q + alpha * (reward - Q)
Q
```

Run the update in your head: if we keep winning, $Q$ climbs toward 1.

::: {.callout-note}
This little rule is famous. It is called the **Rescorla–Wagner rule**, and
brain cells that release dopamine really do signal prediction errors —
this model helped scientists discover that!
:::

## Idea 2: turn guesses into choices

If Machine B looks better, should the agent *always* choose B?
Then it would never notice if it was wrong about A.
The agent should *mostly* choose what looks best, but sometimes explore.

The standard tool is the **softmax** rule. Don't worry about the formula —
what matters is the behaviour of its parameter $\beta$ (beta):

```{r}
softmax <- function(values, beta) {
  scores <- exp(beta * (values - max(values)))
  scores / sum(scores)
}

Q <- c(0.3, 0.5)      # machine B looks a bit better
softmax(Q, beta = 0)   # beta = 0: ignore the values, choose 50:50
softmax(Q, beta = 5)   # beta = 5: clearly prefer B, still try A sometimes
softmax(Q, beta = 50)  # beta = 50: almost always B
```

Small $\beta$ = curious and random. Big $\beta$ = greedy and confident.

## Building the full agent

Now we connect the loop: **look → choose → get reward → learn**, repeated.

```{r}
simulate_learner <- function(n_trials = 100,
                             alpha = 0.3,
                             beta = 5,
                             reward_probs = c(0.3, 0.7)) {
  n_options <- length(reward_probs)
  Q <- rep(0, n_options)
  choice <- integer(n_trials)
  reward <- integer(n_trials)

  for (t in 1:n_trials) {
    p <- softmax(Q, beta)                                   # how attractive is each option?
    choice[t] <- sample.int(n_options, size = 1, prob = p)  # choose one
    reward[t] <- rbinom(1, size = 1, prob = reward_probs[choice[t]])  # world responds
    Q[choice[t]] <- Q[choice[t]] + alpha * (reward[t] - Q[choice[t]]) # learn
  }

  data.frame(trial = 1:n_trials, choice = choice, reward = reward)
}
```

That's it. **A learning mind in 15 lines.** Let's watch it live its life:

```{r}
library(ggplot2)
set.seed(1)

one_life <- simulate_learner()
ggplot(one_life, aes(x = trial, y = choice)) +
  geom_point(aes(color = factor(reward)), size = 2) +
  scale_y_continuous(breaks = c(1, 2), labels = c("A (30%)", "B (70%)")) +
  scale_color_manual(values = c("grey70", "gold2"), name = "reward",
                     labels = c("nothing", "coin!")) +
  labs(title = "One agent's 100 turns", y = "machine chosen")
```

At the start it jumps around (exploring); later it settles on B (exploiting).

## The learning curve

One agent is noisy. Let's average **1000 agents** to see the typical pattern:

```{r}
many <- do.call(rbind, lapply(1:1000, function(i) simulate_learner()))
curve <- aggregate(chose_best ~ trial,
                   data = transform(many, chose_best = as.numeric(choice == 2)),
                   FUN = mean)

ggplot(curve, aes(trial, chose_best)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  ylim(0, 1) +
  labs(title = "Learning curve of 1000 agents",
       y = "fraction choosing the better machine")
```

This shape — fast improvement, then a plateau — appears in real experiments
with humans, monkeys, pigeons, bees, and fish. Our 15-line agent captures
something real.

::: {.callout-tip}
## Try it! — become a parameter scientist

Change one parameter at a time and re-run the learning-curve code.
Before each run, **write down your prediction**!

1. `alpha = 0.05` (very slow learner) — what happens to the curve?
2. `alpha = 1` (only the last experience counts) — better or worse? Why?
3. `beta = 0` — what does the curve look like now?
4. Make the problem *hard*: `reward_probs = c(0.45, 0.55)`. How well do
   agents do? How is this like real life?
:::

::: {.callout-important}
## The big idea

$\alpha$ and $\beta$ are called **parameters**. Different parameter values =
different "personalities" of learner. When researchers fit these parameters
to real people's choices, they can *measure* how individuals learn.
That is **computational modelling of behaviour** — the field this camp
is about, and the topic of active research at RIKEN.
:::

## What you learned

- The bandit problem: explore vs exploit
- Agents learn by **prediction error** (Rescorla–Wagner rule), scaled by learning rate $\alpha$
- Softmax turns value estimates into choices; $\beta$ sets curiosity vs greed
- Averaging many simulated agents reveals the *typical* behaviour of a model

These functions are saved in [`R/bandit_functions.R`](../R/bandit_functions.R),
so next time we can just load them with `source("R/bandit_functions.R")`.

Next tutorial: our agent stops being alone. **What happens when agents
can watch each other?**
