---
title: "Tutorial 4: Learning From Others"
subtitle: "Social learning strategies and the power of conformity"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

::: {.callout-note}
## Word list（単語リスト）

| English | 日本語 | Meaning |
|---|---|---|
| social learning | 社会学習 | learning by observing others |
| individual learning | 個人学習 | learning from your own experience |
| conformity | 同調 | following the majority more than proportionally |
| strategy | 戦略 | a rule for deciding what to do |
| majority | 多数派 | the option most people chose |
:::

## Why copy others?

You arrive in a new town, hungry. Two ramen shops: one has a queue,
one is empty. Where do you go?

Most people join the queue — and that is often *smart*. Other people's
choices contain **information**: they may know something you don't.
Humans are the champions of **social learning**; it is the foundation
of culture, science, and TikTok trends alike.

But copying has a dark side, which we will discover in Tutorial 5.
First, let's give our agents the ability to watch each other.

## Setup

Load the functions we built last time (they live in `R/bandit_functions.R`):

```{r}
source("R/bandit_functions.R")
library(ggplot2)
set.seed(2026)
```

## How should you use the majority?

Suppose 10 agents played last round: **7 chose B, 3 chose A**.
You get to peek before choosing. How strongly should the majority pull you?

One flexible rule: choose option $k$ with probability proportional to
(number of people choosing $k$) raised to a power $\theta$ (theta):

$$
p(\text{copy option } k) \propto (\text{count}_k)^{\,\theta}
$$

```{r}
social_weights   # look at the code (it's tiny)
```

The exponent $\theta$ is the personality of the copier:

```{r}
counts <- c(3, 7)   # A was chosen 3 times, B was chosen 7 times
social_weights(counts, theta = 0)   # ignore popularity completely
social_weights(counts, theta = 1)   # copy in proportion: 30% / 70%
social_weights(counts, theta = 3)   # CONFORMITY: majority is exaggerated!
```

With $\theta = 3$, a 70% majority attracts about 92% of copiers.
Following the majority *more than proportionally* is called
**conformist transmission**, and experiments show humans (and even fish
and bees) often do exactly this.

```{r}
#| code-fold: true
#| fig-height: 3.5
grid <- expand.grid(share = seq(0.05, 0.95, by = 0.05),
                    theta = c(0.5, 1, 3))
grid$p_copy <- mapply(function(s, th) social_weights(c(10 * s, 10 * (1 - s)), th)[1],
                      grid$share, grid$theta)
ggplot(grid, aes(share, p_copy, color = factor(theta))) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  labs(title = "How theta changes copying",
       x = "share of group choosing option A",
       y = "probability of copying A",
       color = expression(theta)) +
  coord_equal()
```

The dotted line is proportional copying. Above it (steep S-shape, $\theta=3$):
conformity. Below it ($\theta=0.5$): the majority is *weakened*.

## A society of learners

Now the main event. We put **many agents in the same world**. Each agent
mixes two sources of information:

- its **own experience** (the Q values from Tutorial 3), and
- **what the group did** on the previous round (the social weights above).

One number, $\sigma$ (sigma, from 0 to 1), sets the mix:

$$
p(\text{choice}) = (1 - \sigma) \times p_{\text{individual}} \;+\; \sigma \times p_{\text{social}}
$$

- $\sigma = 0$: pure individual learner — ignores everybody
- $\sigma = 1$: pure copier — never uses its own experience
- in between: a bit of both (like most of us!)

The function `simulate_group()` in `R/bandit_functions.R` implements this.
Open the file and read it — you already know every ingredient!

```{r}
sim <- simulate_group(n_agents = 10, n_trials = 100,
                      sigma = 0.3, theta = 2)
head(sim)
```

Each row = one agent on one trial. Let's watch the whole group at once:

```{r}
ggplot(sim, aes(x = trial, y = agent, fill = factor(choice))) +
  geom_tile() +
  scale_fill_manual(values = c("1" = "grey80", "2" = "seagreen"),
                    name = "choice", labels = c("A (30%)", "B (70%)")) +
  scale_y_continuous(breaks = 1:10) +
  labs(title = "10 agents learning together (green = better option)")
```

Green spreads through the group like a rumour. That is culture, in miniature.

## Does copying help?

Let's compare societies with different amounts of social learning.
`average_performance()` runs the simulation many times and reports,
for each trial, the fraction of agents choosing the better option:

```{r}
#| cache: true
mix <- do.call(rbind, lapply(c(0, 0.3, 0.6), function(s) {
  perf <- average_performance(n_reps = 30, n_agents = 10, n_trials = 100,
                              sigma = s, theta = 2)
  transform(perf, sigma = s)
}))

ggplot(mix, aes(trial, correct, color = factor(sigma))) +
  geom_line(linewidth = 1) +
  ylim(0, 1) +
  labs(title = "Is a copying society a smarter society?",
       y = "fraction choosing the better option",
       color = expression(sigma))
```

Look carefully: societies with moderate copying ($\sigma = 0.3$–$0.6$)
typically end up **better than pure individual learners**. Copying spreads
good discoveries quickly, and the majority filters out individual mistakes.
Group intelligence for free!

...or is it? (Cliffhanger for Tutorial 5.)

::: {.callout-tip}
## Try it! — explore the society

1. **Big city vs small village**: run the comparison with
   `n_agents = 3` and `n_agents = 30`. Does group size change how much
   copying helps?
2. **The contrarian society**: set `theta = 0.5` (avoid the majority).
   How does it perform?
3. **The pure-copier trap**: try `sigma = 0.95`. Watch the tile plot from
   earlier — what goes wrong? Can a group learn if almost nobody collects
   their own information?
:::

::: {.callout-important}
## The big idea

$\sigma$ (how much to copy) and $\theta$ (how strongly to follow majorities)
are **social learning strategies**. Evolution, development, and culture all
tune these knobs. A central question in current research — including at
RIKEN and in the COSMOS summer school — is: *which strategies are smart,
and when?* Tomorrow you will answer part of that question yourselves.
:::

## What you learned

- Other people's behaviour is information; copying is often rational
- $\theta$: proportional imitation vs **conformity** (exaggerating majorities)
- $\sigma$: the balance between own experience and social information
- Moderate copying can make the whole group smarter

Next tutorial: when crowds are wise, when they go mad — and how to tell
the difference.
