---
title: "Tutorial 5: The Wisdom and Madness of Crowds (Interactive Notebook)"
subtitle: "When groups are smarter than individuals — and when they are not"
output:
  html_notebook:
    toc: true
    theme: cosmo
editor_options:
  chunk_output_type: inline
---

::: {.callout-note}
## How to use this notebook

Open `ScienceHighSchool.Rproj` first. Run each code block with the green **Run** button, or place your cursor in the code and press **Cmd/Ctrl + Enter**.

Change one small thing at a time. Then run the block again and look at the result.
:::

```{r setup, include=FALSE}
project_root <- normalizePath("..")
if (!file.exists(file.path(project_root, "ScienceHighSchool.Rproj"))) {
  project_root <- getwd()
}
if (!file.exists(file.path(project_root, "ScienceHighSchool.Rproj"))) {
  stop("Please open ScienceHighSchool.Rproj before running this notebook.")
}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = project_root)
```


::: {.callout-note}
## Word list（単語リスト）

| English | 日本語 | Meaning |
|---|---|---|
| wisdom of crowds | 集団の知恵 | groups being more accurate than individuals |
| herding | 群れ行動・付和雷同 | everyone following everyone, ignoring evidence |
| collective behaviour | 集団行動 | patterns produced by many interacting individuals |
| environment change | 環境変化 | when the best option stops being the best |
| trade-off | トレードオフ | improving one thing makes another worse |
:::

## Part 1: The wisdom of crowds

In 1906, the scientist Francis Galton visited a country fair where ~800
people guessed the weight of an ox. Almost nobody was right. But the
**average of all guesses** missed the true weight by less than 1%.

Why does this work? Let's simulate it. Say the true answer is 500
(beans in a jar, weight of an ox, whatever). Each person's guess is
noisy — sometimes too high, sometimes too low:

```{r}
library(ggplot2)
set.seed(2026)

truth <- 500
guesses <- rnorm(800, mean = truth, sd = 150)  # 800 noisy guesses

ggplot(data.frame(guesses), aes(guesses)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  geom_vline(xintercept = truth, color = "red", linewidth = 1) +
  geom_vline(xintercept = mean(guesses), color = "gold2", linewidth = 1) +
  labs(title = "800 noisy guesses (red = truth, yellow = crowd average)")
```

Individual errors are huge, but they point in *different directions* —
so when we average, they **cancel out**. How big must a crowd be?

```{r}
crowd_error <- function(n_people) {
  mean(replicate(1000, abs(mean(rnorm(n_people, truth, sd = 150)) - truth)))
}
sizes <- c(1, 2, 5, 10, 25, 50, 100, 400)
errors <- sapply(sizes, crowd_error)

ggplot(data.frame(sizes, errors), aes(sizes, errors)) +
  geom_line(color = "steelblue") + geom_point(size = 2) +
  labs(title = "Crowd size vs average error",
       x = "number of people (log scale)", y = "average error") +
  scale_x_log10()
```

::: {.callout-tip}
## Try it! — break the wisdom

The magic needs errors that cancel out. Make everyone share the *same bias*:

```r
guesses <- rnorm(800, mean = truth * 0.7, sd = 150)  # everyone guesses low
```

How wrong is the crowd average now? Can more people fix a shared bias?

*This is why copying is dangerous: it makes people's errors similar.*
:::

## Part 2: When the world changes

In Tutorial 4, our copying societies looked smart. But we tested them in a
**stable world**, where the best option stays best forever.
Real worlds change: shops close, trends die, environments shift.

The machines begin with reward probabilities 0.50 for A and 0.70 for B.
At `change_point = 76`, they secretly **swap**, so A becomes better.
Nobody is told.

```{r}
source("R/bandit_functions.R")
set.seed(2026)

sim <- simulate_group(n_agents = 10, n_trials = 150,
                      reward_probs = c(0.5, 0.7),
                      sigma = 0.7, theta = 4,     # strong social influence
                      alpha = 0.7, beta = 5,
                      change_point = 76)

ggplot(sim, aes(trial, agent, fill = factor(choice))) +
  geom_tile() +
  geom_vline(xintercept = 75.5, color = "red", linewidth = 1) +
  scale_fill_manual(values = c("1" = "grey80", "2" = "seagreen"),
                    name = "choice") +
  scale_y_continuous(breaks = 1:10) +
  labs(title = "Heavy copiers in a changing world (world flips at red line)")
```

This example combines a fast learning rate ($\alpha = 0.7$), much social
learning ($\sigma = 0.7$), and strong conformity ($\theta = 4$). The group
can coordinate quickly, but it can also become rigid.

Look at what happens after the red line. The group often **keeps choosing
the old option** — everyone copies everyone, and the majority feeds on
itself. This is **herding**. (If your plot shows the group escaping,
run it again a few times — sometimes they escape, often they get stuck.)

## Part 3: The full picture

Let's measure this properly: for several values of $\sigma$, average
many simulations, in a world that flips halfway.

```{r, cache=TRUE}
sweep <- do.call(rbind, lapply(c(0, 0.3, 0.6, 0.7), function(s) {
  perf <- average_performance(n_reps = 40, n_agents = 10, n_trials = 150,
                              reward_probs = c(0.5, 0.7),
                              alpha = 0.7, beta = 5,
                              sigma = s, theta = 4, change_point = 76)
  transform(perf, sigma = s)
}))

ggplot(sweep, aes(trial, correct, color = factor(sigma))) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 75.5, linetype = "dashed") +
  ylim(0, 1) +
  labs(title = "The wisdom AND madness of crowds",
       y = "fraction choosing the (currently) best option",
       color = expression(sigma))
```

Read this plot slowly — it is the punchline of the whole camp:

- **Before the change**: moderate copiers ($\sigma = 0.3, 0.6$) usually beat
  individual learners ($\sigma = 0$). *Wisdom of crowds.*
- **Just after the change**: stronger social learners ($\sigma = 0.6, 0.7$)
  fall hardest. The old majority is now wrong. *Madness of crowds.*
- **Later**: some groups recover and some remain stuck. Here, $\sigma = 0.3$
  gives a useful balance between social information and flexibility.

In this experiment, only $\sigma$ changes. We hold $\alpha = 0.7$,
$\beta = 5$, and $\theta = 4$ constant. The best strategy depends on both
the environment and when we measure success.

::: {.callout-note}
Researchers at RIKEN and elsewhere have tested exactly this with real
people playing bandit games in groups online — and found the same pattern:
conformist copying produces both collective wisdom and, in changing
environments, herding. See Toyokawa, Whalen & Laland (2019),
*Nature Human Behaviour* — ask your instructors about it!
:::

::: {.callout-tip}
## Try it! — questions worth investigating

Each of these could become part of your **final presentation**:
Choose one, then write your prediction before changing the code.

1. **Best strategy in a restless world**: make the world flip every 50
   trials (`change_point` currently supports one flip — can you modify
   `simulate_group()` for repeated flips? Ask us for hints!).
   What $\sigma$ wins now?
2. **Group size**: is herding worse in bigger groups? Try
   `n_agents = 5, 20, 50` with $\sigma = 0.6$.
3. **Conformity strength**: fix $\sigma = 0.5$ and sweep
   `theta = 1, 2, 4`. Is strong conformity ever good?
4. **Individual differences**: give agents different $\alpha$, $\beta$,
   $\sigma$, or $\theta$ values. What happens if two agents are strong
   conformists while the others balance individual and social learning?
   See the optional appendix in Tutorial 4 for starter code.
:::

## Part 4: From models to real behaviour — you are the data

Models are only half of science. The other half: **data from real people**.
Conveniently, four real people are attending this camp.

The game has three machines and 40 turns. Each machine has a hidden chance
of paying one coin. Your goal is to earn as many coins as possible.

### How to play

1. Open `ScienceHighSchool.Rproj` in RStudio.
2. Load the game in the **Console**:

```r
source("R/play_bandit_game.R")
```

3. Everyone plays once without social information. Use a short nickname:

```r
play_game("YourName")
```

4. Type `1`, `2`, or `3` on each turn, then press **Enter**. R tells you
   whether you won a coin and shows your total score.
5. After everyone finishes the solo game, play again with hints:

```r
play_game("YourName", hints = TRUE)
```

### What does a hint mean?

A hint comes from **saved human games**, not simulated agents. On turn 12,
for example, the program looks at turn 12 in all CSV files already stored in
`data/game_records/`. It shows the machine chosen most often. It does not show
other players' rewards or the true reward probabilities.

The hint pool is loaded when your game starts. By default, it includes every
saved solo and social game. This means that play order matters. It also means
that test records should be removed before the real social games begin.

### Where is your data stored?

Each game creates one CSV file in `data/game_records/`, such as
`YourName_solo.csv` or `YourName_social.csv`. Each row stores the player,
mode, trial, choice, reward, and hint shown. Replaying with the same nickname
and mode replaces the old file.

These files may contain names. Keep them local. Do not upload or commit them.

To inspect the filenames, run:

```r
list_game_records()
```

To delete an exact test record, use the same nickname and mode:

```r
delete_game_record("Test", hints = FALSE) # delete Test_solo.csv
delete_game_record("Test", hints = TRUE)  # delete Test_social.csv
```

Deletion cannot be undone. Check the filename first, and do not delete another
student's record.

### Analyse the group

After everyone finishes, compare your learning curve with the model's:

```{r, eval=FALSE}
files <- list.files("data/game_records", pattern = "\\.csv$", full.names = TRUE)
human <- do.call(rbind, lapply(files, read.csv))

# machine 2 pays best (0.65) - how often did humans find it?
human_curve <- aggregate(found_best ~ trial,
                         data = transform(human, found_best = as.numeric(choice == 2)),
                         FUN = mean)
plot(human_curve, type = "l", ylim = c(0, 1),
     main = "Did WE find the best machine?")
```

Questions you can now investigate like a real cognitive scientist:

- Do humans learn faster or slower than the model? What $\alpha$ and $\beta$
  make the model *most human-like*?
- Did players with social hints do better than solo players?
- Did anyone show conformity — following hints even after bad experiences?

## What you learned

- Independent errors cancel: crowds can be wiser than any individual
- Copying makes errors *dependent* — herding in changing worlds
- The same strategy can be smart or disastrous depending on the environment
- Models become human behavioural science when compared with real behaviour — including yours

**Now: pick a question, run simulations, make figures, and build your
presentation.** See [`project/mini_project_menu.md`](../project/mini_project_menu.md).
