---
title: "Tutorial 5: The Wisdom and Madness of Crowds (Interactive Notebook)"
subtitle: "When groups are smarter than individuals — and when they are not"
output:
  html_notebook:
    toc: true
    theme: cosmo
editor_options:
  chunk_output_type: inline
---

::: {.callout-note}
## How to use this notebook

Open `ScienceHighSchool.Rproj` first. Run each code block with the green **Run** button, or place your cursor in the code and press **Cmd/Ctrl + Enter**.

Change one small thing at a time. Then run the block again and look at the result.
:::

```{r setup, include=FALSE}
project_root <- normalizePath("..")
if (!file.exists(file.path(project_root, "ScienceHighSchool.Rproj"))) {
  project_root <- getwd()
}
if (!file.exists(file.path(project_root, "ScienceHighSchool.Rproj"))) {
  stop("Please open ScienceHighSchool.Rproj before running this notebook.")
}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = project_root)
```


::: {.callout-note}
## Word list（単語リスト）

| English | 日本語 | Meaning |
|---|---|---|
| wisdom of crowds | 集団の知恵 | groups being more accurate than individuals |
| herding | 群れ行動・付和雷同 | everyone following everyone, ignoring evidence |
| collective behaviour | 集団行動 | patterns produced by many interacting individuals |
| environment change | 環境変化 | when the best option stops being the best |
| trade-off | トレードオフ | improving one thing makes another worse |
:::

## Part 1: The wisdom of crowds

In 1906, the scientist Francis Galton visited a country fair where ~800
people guessed the weight of an ox. Almost nobody was right. But the
**average of all guesses** missed the true weight by less than 1%.

Why does this work? Let's simulate it. Say the true answer is 500
(beans in a jar, weight of an ox, whatever). Each person's guess is
noisy — sometimes too high, sometimes too low:

```{r}
library(ggplot2)
set.seed(2026)

truth <- 500
guesses <- rnorm(800, mean = truth, sd = 150)  # 800 noisy guesses

ggplot(data.frame(guesses), aes(guesses)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  geom_vline(xintercept = truth, color = "red", linewidth = 1) +
  geom_vline(xintercept = mean(guesses), color = "gold2", linewidth = 1) +
  labs(title = "800 noisy guesses (red = truth, yellow = crowd average)")
```

Individual errors are huge, but they point in *different directions* —
so when we average, they **cancel out**. How big must a crowd be?

```{r}
crowd_error <- function(n_people) {
  mean(replicate(1000, abs(mean(rnorm(n_people, truth, sd = 150)) - truth)))
}
sizes <- c(1, 2, 5, 10, 25, 50, 100, 400)
errors <- sapply(sizes, crowd_error)

ggplot(data.frame(sizes, errors), aes(sizes, errors)) +
  geom_line(color = "steelblue") + geom_point(size = 2) +
  labs(title = "Crowd size vs average error",
       x = "number of people (log scale)", y = "average error") +
  scale_x_log10()
```

::: {.callout-tip}
## Try it! — break the wisdom

The magic needs errors that cancel out. Make everyone share the *same bias*:

```r
guesses <- rnorm(800, mean = truth * 0.7, sd = 150)  # everyone guesses low
```

How wrong is the crowd average now? Can more people fix a shared bias?

*This is why copying is dangerous: it makes people's errors similar.*
:::

## Part 2: When the world changes

Yesterday our copying societies looked smart. But we tested them in a
**stable world**, where the best option stays best forever.
Real worlds change: shops close, trends die, environments shift.

`simulate_group()` can change the world in the middle:
at `change_point = 76`, machines secretly **swap**
(A becomes the good one!). Nobody is told.

```{r}
source("R/bandit_functions.R")
set.seed(2026)

sim <- simulate_group(n_agents = 10, n_trials = 150,
                      sigma = 0.9, theta = 2,     # a society of heavy copiers
                      change_point = 76)

ggplot(sim, aes(trial, agent, fill = factor(choice))) +
  geom_tile() +
  geom_vline(xintercept = 75.5, color = "red", linewidth = 1) +
  scale_fill_manual(values = c("1" = "grey80", "2" = "seagreen"),
                    name = "choice") +
  scale_y_continuous(breaks = 1:10) +
  labs(title = "Heavy copiers in a changing world (world flips at red line)")
```

Look at what happens after the red line. The group often **keeps choosing
the old option** — everyone copies everyone, and the majority feeds on
itself. This is **herding**. (If your plot shows the group escaping,
run it again a few times — sometimes they escape, often they get stuck.)

## Part 3: The full picture

Let's measure this properly: for several values of $\sigma$, average
many simulations, in a world that flips halfway.

```{r, cache=TRUE}
sweep <- do.call(rbind, lapply(c(0, 0.3, 0.6, 0.9), function(s) {
  perf <- average_performance(n_reps = 40, n_agents = 10, n_trials = 150,
                              sigma = s, theta = 2, change_point = 76)
  transform(perf, sigma = s)
}))

ggplot(sweep, aes(trial, correct, color = factor(sigma))) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 75.5, linetype = "dashed") +
  ylim(0, 1) +
  labs(title = "The wisdom AND madness of crowds",
       y = "fraction choosing the (currently) best option",
       color = expression(sigma))
```

Read this plot slowly — it is the punchline of the whole camp:

- **Before the change**: copiers ($\sigma = 0.3, 0.6$) beat individual
  learners ($\sigma = 0$). *Wisdom of crowds.*
- **After the change**: heavy copiers ($\sigma = 0.9$) crash and often never
  recover. *Madness of crowds.* Individual learners adapt fastest.
- **Moderate copiers** get most of the wisdom with little of the madness.

One parameter, $\sigma$, moves a society between collective intelligence
and collective stupidity. Fashion bubbles, stock-market crashes, and
misinformation cascades all rhyme with this simple model.

::: {.callout-note}
Researchers at RIKEN and elsewhere have tested exactly this with real
people playing bandit games in groups online — and found the same pattern:
conformist copying produces both collective wisdom and, in changing
environments, herding. See Toyokawa, Whalen & Laland (2019),
*Nature Human Behaviour* — ask your instructors about it!
:::

::: {.callout-tip}
## Try it! — questions worth investigating

Each of these could become part of your **final presentation**:

1. **Best strategy in a restless world**: make the world flip every 50
   trials (`change_point` currently supports one flip — can you modify
   `simulate_group()` for repeated flips? Ask us for hints!).
   What $\sigma$ wins now?
2. **Group size**: is herding worse in bigger groups? Try
   `n_agents = 5, 20, 50` with $\sigma = 0.6$.
3. **Conformity strength**: fix $\sigma = 0.5$ and sweep
   `theta = 1, 2, 4`. Is strong conformity ever good?
:::

## Part 4: From models to real behaviour — you are the data

Models are only half of science. The other half: **data from real people**.
Conveniently, four real people are attending this camp.

Play the slot-machine game (each person, solo first, then with hints):

```r
source("R/play_bandit_game.R")
play_game("YourName")               # solo
play_game("YourName", hints = TRUE) # with social information
```

Your choices are saved in `data/game_records/`. Then compare **your**
learning curve with the model's:

```{r, eval=FALSE}
files <- list.files("data/game_records", full.names = TRUE)
human <- do.call(rbind, lapply(files, read.csv))

# machine 2 pays best (0.65) - how often did humans find it?
human_curve <- aggregate(found_best ~ trial,
                         data = transform(human, found_best = as.numeric(choice == 2)),
                         FUN = mean)
plot(human_curve, type = "l", ylim = c(0, 1),
     main = "Did WE find the best machine?")
```

Questions you can now investigate like a real cognitive scientist:

- Do humans learn faster or slower than the model? What $\alpha$ and $\beta$
  make the model *most human-like*?
- Did players with social hints do better than solo players?
- Did anyone show conformity — following hints even after bad experiences?

## What you learned

- Independent errors cancel: crowds can be wiser than any individual
- Copying makes errors *dependent* — herding in changing worlds
- The same strategy can be smart or disastrous depending on the environment
- Models become science when compared with real behaviour — including yours

**Now: pick a question, run simulations, make figures, and build your
presentation.** See [`project/mini_project_menu.md`](../project/mini_project_menu.md).
